---
title: "Measuring contraception"
description: |
  How we measured methods of contraception, reasons for using different contraceptives, and fertility awareness.
author:
  - name: "Ruben C. Arslan"
    url: https://rubenarslan.github.io
    affiliation_url: https://www.mpib-berlin.mpg.de/en/staff/ruben-arslan
    affiliation: "Center for Adaptive Rationality, Max Planck Institute for Human Development, Berlin" 
journal: 
  title: "Open Science Framework"
doi: "10.17605/OSF.IO/3DKYS"
slug: arslanetal2019contraception
citation_url: https://osf.io/3dkys/
date: 04-02-2019
categories: 
  - sex diary
  - open science
  - contraception
output:
  radix::radix_article:
    self_contained: false
draft: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

In our sex diary study, one of the most important components was our contraception
questionnaire. Surprisingly, to us, we did not find a good, open standard questionnaire
that fit our needs.^[We checked a few familiar demographic surveys, which erred on the side of asking fewer questions, and the psychological literature, which erred on the side of not reporting exactly which questions were asked] We needed something that would measure

- not only one main, but all methods of contraception
- fertility awareness^[Counting days, measuring temperature, mucus] independently of whether it was used for contraceptive purposes, i.e. including apps like [Clue](https://helloclue.com/)
- the reasons why people mix contraceptive methods
- the exact contraceptive pill, intra-uterine device, ring, patch etc.^[so we could differentiate between contraceptives with different kinds and amounts of progestogens and different amounts of estrogens]
- what fertility-awareness method users do when fertile
- all potential contraceptive methods that we could imagine

Why do we need such detailed information on contraception to study ovulatory effects?

1. Combined hormonal contraceptives suppress ovulation—we could use them as a quasi-control group for [our last study](https://psycnet.apa.org/record/2018-41799-001). However, progestogen-only contraceptives do not reliably suppress ovulation.
2. Some fertility awareness method users are sexually abstinent in the fertile window. When we are interested in whether ovulation whets the appetite for sex and changes sexual behaviour, we should probably exclude these women or at least treat them separately.
3. Awareness of the cycle phase could lead to different responses. Potentially, women actually feel or report feeling more desirable, or more sexual when they _believe_ they are fertile—independently of, or on top of any hormonally caused changes.^[In our last study, we only found some descriptive differences between non-hormonal methods. The only significant difference was between hormonal and non-hormonal methods, but our sample size for fertility awareness methods was small and potentially women who used condoms also used a fertility tracking app for fun, but not contraception. Descriptively, fertile window increases in extra-pair desire [were larger for women using awareness methods than for barrier method (condoms, mostly) users](https://rubenarslan.github.io/ovulatory_shifts/3_fertility_robustness.html#m_m1-moderation-by-contraceptive-method), but [the opposite pattern held for in-pair desire](https://rubenarslan.github.io/ovulatory_shifts/3_fertility_robustness.html#m_m1-moderation-by-contraceptive-method-3), and we found [no apparent difference for self-perceived desirability](https://rubenarslan.github.io/ovulatory_shifts/3_fertility_robustness.html#m_m1-moderation-by-contraceptive-method-4).]

Today, I want to show the questions we chose and share the questionnaire. We're 
still in the market for improvements. At the same time, I want to encourage others to more fully
report contraception information in their menstrual cycle studies. When you are reporting effects on
sexual desire or activity, we need to know more than that participants were not on the pill and not pregnant. I'll try to show why in the coming blog posts.


<details>
<summary>Translation details</summary>

To be able to show the descriptive graphs below, we translated the questions.^[If you ever find yourself in the situation of wanting to re-label items in your language with English labels, you can follow along the code for this post on Github. It's really simple as long as you have a well-structured item sheet.]

```{r}
library(formr)
library(codebook)
library(tidyverse)
theme_set(theme_bw())
load("~/research/routine_and_sex/cleaned.rdata")
questions <- rio::import("https://docs.google.com/spreadsheets/d/1tLQDVyYUAXLBkblTT8BXow_rcg5G6xK9Vi3xTGieN20/edit#gid=0", format = "xlsx")
choices <- rio::import("https://docs.google.com/spreadsheets/d/1tLQDVyYUAXLBkblTT8BXow_rcg5G6xK9Vi3xTGieN20/edit#gid=1116762580", which = 2)
questions <- questions %>% 
  slice(43:72) %>% 
  mutate_if(is.character, ~ if_na(., "")) %>% select(type:label_en)

choices <- choices %>% 
  slice(1:182) %>% 
  filter(!is.na(name), name != "") %>% 
  mutate(
    list_name = na_if(list_name, ""),
    list_name = zoo::na.locf(list_name)
    )

# loop over all questions in the sheet
for (i in seq_along(questions$name)) {
  var <- questions$name[i]
  # replace the variable label
  if (exists(var, all_surveys)) {
    var_label(all_surveys[[var]]) <- questions %>% filter(name == var) %>% pull(label_en)
    choice_list_name <- questions %>% filter(name == var) %>% 
      select(type) %>%
      separate(type, c("type", "list_name"), sep = " ", fill = "right") %>% 
      pull(list_name)
    # find out if there were labelled choices
    translated_choices <- choices %>% 
          filter(list_name == choice_list_name) %>% 
          select(label_en, name) %>% 
          mutate(name = type.convert(name, as.is = TRUE)) %>% 
          mutate_if(is.numeric, as.double) %>% 
      # make sure they're the right type
          deframe()
    if ( !is.null(val_labels(all_surveys[[var]]) ) && length(translated_choices) > 0) {
    # if so overwrite
      val_labels(all_surveys[[var]]) <- translated_choices
    }
  }
}

```

</details>

## Contraception or not?

We began by asking whether women used contraception at all. This time, we only asked
women who were not pregnant. Next time, we plan to ask all women about contraception
because even pregnant women could e.g. use condoms to prevent sexually transmitted infections,
urinary tract infections. However, so that it does not like a lack of forethought, 
we should probably optionally display some explanation to this effect for e.g. menopausal and pregnant women, because some women found it odd to be asked these questions. In this study, we gave
women a free text field to write in if they wanted to specify other reasons why they 
didn't use contraception. 

```{r fig.cap="Pregnant women were excluded from the contraception questionnaire."}
plot_labelled(all_surveys$pregnant) + 
    geom_text(aes(y = ..count.., label = ..count..), 
              stat = "count", 
              color = "#4F94B0",
              vjust = -0.25, 
              angle = 0, na.rm = TRUE) +
  ggtitle("Pregnant")
```

```{r fig.cap="Different degrees of using contraception", fig.height=7}
contraception_at_all <- haven::as_factor(na.omit(all_surveys$contraception_at_all), levels = "labels")
plot_labelled(contraception_at_all) + 
    geom_text(aes(y = ..count.., label = ..count..), 
              stat = "count", 
              color = "#4F94B0",
              vjust = -0.25, 
              angle = 0, na.rm = TRUE) +
  ggtitle("Contraception at all?")
```


## Methods

Next, we asked all women who used contraception which methods they used. Multiple
methods could be checked. As you can see, condoms and the the pill were the two most
common methods. A lot of women also combined both. Next, you might be surprised at the
number of women who report using _coitus interruptus_ (in the lower left graph), 
the pull-out method. However, more women combined this with condoms or a fertility
awareness method than used only _coitus interruptus_. 
Next were intra-uterine devices and hormonal contraceptives other than the pill.
We will look at these in detail next.

```{r fig.width=10, fig.height=9, layout = "l-body-outset", fig.cap = "Methods of contraception and how they are combined"}
library(UpSetR)
comma_separated_to_columns <- function(df, col) {
  colname <- deparse(substitute(col))
  df$splitcol <- df %>% pull(colname)
  separate_rows(df, splitcol, convert = TRUE, sep = ", ") %>% 
    mutate(splitcol = if_else(is.na(splitcol), "no", 
                        if_else(splitcol == "" | 
                                  splitcol %in% c(), "other", as.character(splitcol)))) %>% 
    mutate(#splitcol = stringr::str_c(colname, "_", splitcol), 
           value = 1) %>% 
    spread(splitcol, value, fill = 0) %>% 
    select(-colname)
}
library(UpSetR)
all_surveys %>% select(session, contraception_method) %>% 
  filter(!is.na(contraception_method), contraception_method != "") %>% 
  comma_separated_to_columns(contraception_method) %>% 
  select(-session) %>% 
  {
  upset(., ncol(.), 20, show.numbers = TRUE, order.by = "freq",
      main.bar.color = "#6E8691",
      matrix.color = "#6E8691",
      sets.bar.color = "#4F94B0")
  }
```


As you can see, the non-pill hormonal methods were mainly NuvaRing (a estrogen/progestogen vaginal ring) and Mirena (a progestogen-only intra-uterine contraceptive).

```{r fig.cap="Other hormonal contraceptives"}
plot_labelled(na.omit(all_surveys$contraception_hormonal_other)) + 
    geom_text(aes(y = ..count.., label = ..count..), 
              stat = "count", 
              color = "#4F94B0",
              vjust = -0.25, 
              angle = 0, na.rm = TRUE) +
  ggtitle("Other hormonal contraceptives")
```

## Combining methods

Above, we saw that many women combine multiple contraceptives. We had a list of reasons
that we considered likely, but many women also gave reasons we hadn't included.

```{r fig.width=10, fig.height=8, layout = "l-body-outset", fig.cap = "Why combine several contraceptives?"}
all_surveys %>% select(session, contraception_combi) %>% 
  filter(!is.na(contraception_combi), contraception_combi != "") %>% 
  comma_separated_to_columns(contraception_combi) %>% 
  select(-session) %>% 
  {
  upset(., ncol(.), 20, show.numbers = TRUE, order.by = "freq",
      main.bar.color = "#6E8691",
      matrix.color = "#6E8691",
      sets.bar.color = "#4F94B0")
  }
```

The most common reason was as a fallback method during fertile days (i.e., for women using
fertility awareness methods). Many used condoms in addition to the pill to also prevent
STDs, but many also used the pill and condoms to decrease conception risk even further.
Fallback methods were also common (e.g. condoms as a fallback to the pill, or pull-out
as a fallback to condoms). A substantial number of women also used different methods for
different partners, and a small minority also said they used condoms when their partner
has a cold.

From the reasons we hadn't included in our list, we should include at least the following
in the future:^[Taken from the written responses to the other reasons prompt.]

- fallback in case the pill might not have worked (e.g., stomach flu, throwing up, travel, medication)
- pill against acne/hypermenorrhoe/menstrual pain/other reasons, another method for contraception
- condoms for hygienic reasons
- condoms to avoid urinary tract infections
- according to mood

### Fertility awareness

A special case of combining contraceptive methods are the fertility awareness methods.
How many of our participants are sexually abstinent during the fertile phase, how many switch
to other methods? As you can see, most switch to other methods. Total sexual abstinence is rare, although
quite a few report having penetrative sex less often, or not at all.

```{r fig.width=10, fig.height=8, layout = "l-body-outset", fig.cap = "What do fertility awareness method users do during their fertile phase?"}
all_surveys %>% select(session, contraception_calendar_abstinence) %>% 
  filter(!is.na(contraception_calendar_abstinence), contraception_calendar_abstinence != "") %>% 
  comma_separated_to_columns(contraception_calendar_abstinence) %>% 
  select(-session) %>% 
  {
  upset(., ncol(.), 20, show.numbers = TRUE, order.by = "freq",
      main.bar.color = "#6E8691",
      matrix.color = "#6E8691",
      sets.bar.color = "#4F94B0")
  }
```

<details><summary>Methods and reasons for combining methods in two big graphs</summary>

We can of course also merge the reasons for combining contraceptives with the contraceptive information itself.

```{r fig.width=15, fig.height=10, layout = "l-screen-inset", fig.cap = "Which contraceptives are combined for which reasons?"}
all_surveys %>% select(session, contraception_combi) %>% 
  filter(!is.na(contraception_combi), contraception_combi != "") %>% 
  comma_separated_to_columns(contraception_combi) %>% 
  left_join(all_surveys %>% select(session, contraception_method) %>% 
  filter(!is.na(contraception_method), contraception_method != "") %>% 
  comma_separated_to_columns(contraception_method), by = 'session') %>% 
  select(-session) %>% 
  {
  upset(., ncol(.), 20, show.numbers = TRUE, order.by = "freq",
      main.bar.color = "#6E8691",
      matrix.color = "#6E8691",
      sets.bar.color = "#4F94B0")
  }
```


```{r fig.width=15, fig.height=10, layout = "l-screen-inset", fig.cap = "Which contraceptive method do fertility awareness method users switch to during their fertile phase?"}
all_surveys %>% select(session, contraception_calendar_abstinence) %>% 
  filter(!is.na(contraception_calendar_abstinence), contraception_calendar_abstinence != "") %>% 
  comma_separated_to_columns(contraception_calendar_abstinence) %>% 
  left_join(all_surveys %>% select(session, contraception_method) %>% 
  filter(!is.na(contraception_method), contraception_method != "") %>% 
  comma_separated_to_columns(contraception_method), by = 'session') %>% 
  select(-session) %>% 
  {
  upset(., ncol(.), 20, show.numbers = TRUE, order.by = "freq",
      main.bar.color = "#6E8691",
      matrix.color = "#6E8691",
      sets.bar.color = "#4F94B0")
  }
```
</details>

## The pill

To help our participants identify their pill, we used a searchable drop-down that showed the brand name
of the pill together with a picture of the packaging. Most women found their pill in our list; the others
were asked to enter name and details (progestogen type and amount, estrogen amount) into text fields.

```{r fig.cap="Common oral contraceptive pills. Only those used by at least 10 women shown."}
pills <- all_surveys %>% 
  group_by(contraception_hormonal_pill) %>% 
  summarise(count = n()) %>% 
  arrange(count) %>% 
  filter(count > 9, !is.na(contraception_hormonal_pill)) %>% 
  mutate(contraception_hormonal_pill = forcats::fct_inorder(contraception_hormonal_pill))
ggplot(pills, aes(contraception_hormonal_pill, count)) +
  geom_col() + 
  coord_flip() +
    geom_text(aes(y = count, label = count), 
              color = "#4F94B0",
              vjust = -0.25, 
              angle = 270, na.rm = TRUE) +
  ggtitle("Common pills")
```

```{r fig.cap = "Micrograms of estrogen in a pill (unified to average content over 21 days, without pill break)"}
all_surveys %>% 
  drop_na(contraception_hormonal_pill_estrogen) %>% 
  mutate(contraception_hormonal_pill_estrogen = round(contraception_hormonal_pill_estrogen)) %>% 
  group_by(contraception_hormonal_pill_estrogen) %>% 
ggplot(., aes(factor(contraception_hormonal_pill_estrogen))) +
  geom_bar() +
  geom_text(aes(y = ..count.., label = ..count..), 
              stat = "count", 
              color = "#4F94B0",
              vjust = -0.25, 
              angle = 0, na.rm = TRUE) +
  xlab("Micrograms of estrogen in a pill")
```

## Awareness

Quite a few women used some sort of cycle app.

```{r fig.cap="Cycle apps. Only those used by at least 10 women shown."}
plot_labelled(na.omit(all_surveys$contraception_app)) + 
    geom_text(aes(y = ..count.., label = ..count..), 
              stat = "count", 
              color = "#4F94B0",
              vjust = -0.25, 
              angle = 0, na.rm = TRUE) +
  ggtitle("Cycle apps")

contraception_apps <- all_surveys %>% 
  mutate(contraception_app_name = str_trim(str_to_title(contraception_app_name))) %>% 
  group_by(contraception_app_name) %>% 
  summarise(count = n()) %>% 
  arrange(count) %>% 
  filter(count > 9, !is.na(contraception_app_name)) %>% 
  mutate(contraception_app_name = forcats::fct_inorder(contraception_app_name))
ggplot(contraception_apps, aes(contraception_app_name, count)) +
  geom_col() + 
  coord_flip() +
    geom_text(aes(y = count, label = count), 
              color = "#4F94B0",
              vjust = -0.25, 
              angle = 270, na.rm = TRUE) +
  ggtitle("Cycle apps")
```

We looked at all of these apps and coded their purposee. Some of them are just pill reminders, some additionally allow you to track symptoms, whereas others really lead to fertility awareness (i.e., tell you when they estimate you will be fertile).

```{r}
ggplot(all_surveys %>% filter(cycle_awareness_app != "none"), 
       aes(cycle_awareness_app)) +
  geom_bar() + 
  coord_flip() +
    geom_text(aes(y = ..count.., label = ..count..), 
              stat = 'count',
              color = "#4F94B0",
              vjust = -0.25, 
              angle = 270, na.rm = TRUE) +
  ggtitle("Cycle app awareness")
```

We are really glad we asked about apps. As you can see below, the vast majority
of cycle awareness app users do not use them as a primary method of contraception; 
most use condoms and IUDs.

```{r fig.width=10, fig.height=8, layout = "l-body-outset", fig.cap="Cycle awareness app users' contraceptive methods."}
all_surveys %>% 
  filter(cycle_awareness_app == "cycle_phase_aware", 
         !is.na(contraception_method), contraception_method != "") %>% 
  select(session, contraception_method) %>% 
  comma_separated_to_columns(contraception_method) %>% 
  select(-session) %>% 
  {
  upset(., ncol(.), 20, show.numbers = TRUE, order.by = "freq",
      main.bar.color = "#6E8691",
      matrix.color = "#6E8691",
      sets.bar.color = "#4F94B0")
  }
```

## Estrogen and progestogen

As you could see above, our participants used a wide variety of contraceptives. Among them were quite a few progestogen-only contraceptives (hormonal IUDs, such as Mirena, patches and depots, and of course various mini-pills, such as Jaydess, Cyprella, Desirett, Cerazette, etc.). When we add all this together, we see that a substantial minority uses progestogen-only contraceptives.

```{r}
qplot(all_surveys %>% drop_na(estrogen_progestogen) %>% pull(estrogen_progestogen)) +
  xlab("Contraceptives by hormones") +
    geom_text(aes(y = ..count.., label = ..count..), 
              stat = 'count',
              color = "#4F94B0",
              vjust = -0.25, 
              angle = 0, na.rm = TRUE)
```


## Summary
So, these were the main questions from our questionnaire. I didn't get into how
we combined this with questions about the period (menarche/menopause, regularity, cycle length, etc.),
family planning, and so on.

I added the entire list of questions, including skipping logic (which takes some forethought) below. You can export these to Excel and use them directly in [formr.org](https://formr.org). This is already an adapted version of the questionnaire we used and I'm sure it can be improved further.  What questions are you asking? What can we improve in our next study?

## Our questions

```{r layout="l-screen-inset", echo = FALSE}
DT::datatable(questions, escape = T, caption = "Table 1. Survey sheet",
extensions = 'Buttons', rownames = F, options = list(
    dom = 'Bfrtip',
    buttons = c('copy', 'csv', 'excel')
  )
)

DT::datatable(choices, escape = T, caption = "Table 2. Choices sheet",
extensions = 'Buttons', rownames = F, options = list(
    dom = 'Bfrtip',
    buttons = c('copy', 'csv', 'excel')
  )
)
```

## Acknowledgements {.appendix}

Thanks to Tessa Cappelle and Kim Gloystein for help translating the questionnaire. Thanks to Julie Driebe and Iro Eleftheriadou who helped me clean and code the contraception questionnaire data. The questionnaire was designed by our study team (Julie Driebe, Tanja M. Gerlach, Lars Penke, Julia Jünger (now: Julia Stern), Julia Ostner).
