---
title: "HIBAR: Fertility and Conspicuous Consumption"
description: |
  Had I Been a Reviewer. 
author:
  - name: Ruben C. Arslan
    url: https://rubenarslan.github.io
    affiliation: Psychological Research Methods, University Witten/Herdecke
date: 2026-09-11
categories: 
  - HIBAR
  - post-publication review
  - fertility
  - menstrual cycle
  - ERROR
  - quick job
output:
  distill::distill_article:
    self_contained: false
    toc: true
    toc_depth: 3
---


# Severe errors in ["Fertility increases women's desire for conspicuous luxury"](https://doi.org/10.1371/journal.pone.0354126)
A paper by Kim, Durante, Laran, Nikiforidis, & Griskevicius (2026, PLoS One).

A journalist just asked me for my opinion on this recent publication. The paper reports "that ovulation—the time of peak fertility—increases women's desire to purchase conspicuous luxury goods". Having worked on the topic of menstrual cycle research for a few years now, I was skeptical of these results. I fed the paper into Claude Opus 5 and had it look for some of the problems I typically see in menstrual cycle research (poor fertility measurement, poor hormone measures, small samples, uncorrected multiple testing) as well as errors. I also attached the data, which the authors, to their credit, shared on OSF.

Claude found a lot of errors. The paper has three studies. The studies are poorly designed if the goal is to have a realistic chance of finding non-sexuality-related psychological effects of being ovulatory. They ignore or misrepresent several recommendations that Gangestad, I, and others have made to improve such studies. More worryingly, all of these seem to have severe errors, ranging from basic statistical mistakes to citations that don't back up claims. Once corrected, there is no support for the title claims. I verified two claims myself (the incorrect ANOVA in Study 1 and the incorrectly pooled SEs in Study 2). These two consequential mistakes already cross the threshold where I would urgently recommend the authors correct the paper, so I didn't see the need to go deeper. However, they are just the tip of the iceberg and several other consequential errors were found and look plausible to me. Given the scope and number of errors, including unfixable design problems, a retraction might be warranted.

What follows is Claude Opus 5 output with minor rewriting from me. I then had Claude Fable 5.1 re-run every analysis below against the posted data, read the AsPredicted entry and the abstracts of the two key methodological citations, and adjust the text where a claim was overstated or its consequences were unclear. One claim that was technically true but inconsequential (the paper's "low fertility" prediction point lies 0.0015 below a conception probability of zero) was dropped from the list of errors.

```{r setup}
#| code-fold: true
library(haven)
library(sandwich)
library(knitr)
library(ggplot2)

dd <- here::here("_posts/2026-09-11-hibar-fertility-and-conspicuous-consumption")
s1 <- as.data.frame(read_sav(file.path(dd, "Fertility_Data_Study 1.sav")))
s2 <- as.data.frame(read_sav(file.path(dd, "Fertility_Data_Study 2.sav")))
s3 <- as.data.frame(read_sav(file.path(dd, "Fertility_Data_Study 3.sav")))
s1[] <- lapply(s1, as.numeric); s2[] <- lapply(s2, as.numeric); s3[] <- lapply(s3, as.numeric)
posted_ncol <- c(`Study 1` = ncol(s1), `Study 2` = ncol(s2), `Study 3` = ncol(s3))

fmt_p <- function(p) ifelse(p < .0001, "<.0001", formatC(p, format = "f", digits = 4))
theme_set(theme_minimal(base_size = 11))
```

## What this document does

I downloaded the three SPSS files from the paper's OSF deposit (<https://osf.io/c3v49/>)
and the AsPredicted entry for Study 3 (#153,708, <https://aspredicted.org/C2V_Q4Y>, a PDF copy
is in this post's folder), and
re-ran every analysis for which data are posted. The pretests, the SES-adjusted
three-way interaction in Study 2, and the sexual-orientation controls cannot be re-run
from the deposit. Everything else below is reproducible from those three files plus this
document.

The summary is:

1. Study 1's headline result is wrong. Every reported *F* is a test of the effect at
   SES = 0, which is what SPSS GLM gives you when a covariate is left uncentered. SES is
   a 1--5 scale. Correctly centered, the test of Hypothesis 1 is *p* = .18.
2. Study 1's dependent variable is built from two drawings, one of which is missing
   for 17--26% of participants in exactly the two cells that are hypothesis-relevant, and
   for nobody in the other two. Complete-case analysis is null throughout.
3. Study 1's session order was not randomised and is confounded with fertility. The
   interaction is absent in the 26 women who did the low-fertility session first, and
   recoded as session 1 vs session 2 it is at least as strong as it is as a fertility
   effect. The design cannot separate the two.
4. Study 2's conditional effects use a pooled error term across cells whose variances
   differ by a factor of four. The three-way interaction itself survives robust standard
   errors, but the simple slope that supports Hypothesis 1 (fertility raising purchase
   intention for public luxury) becomes *p* ≈ .08, and the only slope that stays
   significant is the negative one in the private condition, which no hypothesis
   predicted.
5. Study 3's preregistered test failed (*p* = .099). The analysis the conclusion rests on
   removes a preregistered exclusion criterion, and cannot be reproduced because the
   posted data contain only the excluded-sample cases.

---

## Study 1

### The reported *F* statistics are tests at SES = 0

The paper analyses logo area in a repeated-measures ANOVA "with SES as a covariate" and
reports four tests. In SPSS `GLM ... /WSFACTOR`, the *Tests of Within-Subjects Effects*
table does not centre covariates. The row for a within-subjects effect therefore
tests the contrast mean at covariate = 0, not the average contrast mean. SES here is the
mean of two 1--5 items; its observed minimum is 1.

Each within-subjects test is algebraically the intercept test in a regression of the
relevant difference score on the covariate, so all four are reproducible in three lines
of R:

```{r s1-helper}
s1$Low_Self_B <- ifelse(is.na(s1$Low_Self_S),
                        s1$Low_Self_BS,
                        2 * s1$Low_Self_BS - s1$Low_Self_S)   # absent from the posted file

ctest <- function(y, cov = NULL, center = TRUE) {
  ok <- !is.na(y); y <- y[ok]
  if (is.null(cov)) m <- lm(y ~ 1) else {
    x <- cov[ok]; if (center) x <- x - mean(x); m <- lm(y ~ x)
  }
  cf <- summary(m)$coefficients["(Intercept)", ]
  Fv <- cf[["t value"]]^2
  data.frame(n = length(y), M = mean(y), F = Fv, df = m$df.residual,
             p = cf[["Pr(>|t|)"]], eta2p = Fv / (Fv + m$df.residual))
}

with(s1, {
  cs <- list(
    `fertility x target`   = (High_Self_BS - High_Other_BS) - (Low_Self_BS - Low_Other_BS),
    `H1: self, high-low`   =  High_Self_BS  - Low_Self_BS,
    `high fert: self-other`=  High_Self_BS  - High_Other_BS,
    `other, high-low`      =  High_Other_BS - Low_Other_BS)
  out <- do.call(rbind, lapply(names(cs), function(nm) {
    a <- ctest(cs[[nm]], SES, center = FALSE)   # SPSS default
    b <- ctest(cs[[nm]], SES, center = TRUE)    # correct
    d <- ctest(cs[[nm]])                        # no covariate
    data.frame(effect = nm,
               `F uncentered` = round(a$F, 2), `p uncentered` = fmt_p(a$p),
               `F centered`   = round(b$F, 2), `p centered`   = fmt_p(b$p),
               `F no cov`     = round(d$F, 2), `p no cov`     = fmt_p(d$p),
               check.names = FALSE)
  }))
  kable(out, caption = "Study 1: the published F values are the uncentered column.")
})
```

Published values for comparison: 16.19, 6.34, 4.59, 7.88. The uncentered column
matches all four (the last is 7.87 here, 7.8746 unrounded, against the paper's 7.88),
while neither the centered nor the covariate-free column comes close. That leaves little
doubt about the provenance: the published tests are evaluated at SES = 0.

What the extrapolation amounts to:

```{r s1-extrapolation}
with(s1, {
  h1 <- High_Self_BS - Low_Self_BS
  ix <- (High_Self_BS - High_Other_BS) - (Low_Self_BS - Low_Other_BS)
  out <- do.call(rbind, lapply(list(`H1: self, high-low` = h1, `fertility x target` = ix),
    function(y) { m <- lm(y ~ SES); data.frame(
      `intercept (SES = 0)` = round(coef(m)[1], 3),
      `slope per SES point` = round(coef(m)[2], 3),
      `p (slope)` = round(summary(m)$coefficients[2, 4], 3),
      `sample mean effect` = round(mean(y), 3), check.names = FALSE) }))
  out <- cbind(effect = c("H1: self, high-low", "fertility x target"), out)
  kable(out, row.names = FALSE,
        caption = sprintf("SES ranges %.1f to %.1f in this sample.", min(SES), max(SES)))
})
```

The published "significant" H1 effect of 0.85 in² is the intercept of a line evaluated
five scale points below the lowest observed SES. The actual average effect is 0.16 in²,
*p* = .18.

So Hypothesis 1 is not supported in Study 1 under any specification I can think of.
The one effect that survives is the interaction (*p* = .017 centered, *p* = .026 without
the covariate), and it is carried by the *decrease* in logo size drawn for other women,
which the paper itself calls unexpected and then treats as confirmation. The sentence
"fertility increased logo size, as women drew significantly bigger logos in the high
fertility ... study session (F(1, 68) = 6.34, p = .014)" is simply not true of these
data. The 0.16 in² difference is not distinguishable from zero.

### The dependent variable is not comparable across cells

The DV is the mean of a handbag logo and a shoe logo. Shoe drawings are missing in a
strongly non-random pattern, and the composite silently falls back to the bag value
alone. `Low_Self_B` is missing from the posted file entirely; I reconstructed it above.

```{r s1-missing}
cells <- c("Low_Self", "High_Self", "Low_Other", "High_Other")
kable(do.call(rbind, lapply(cells, function(cl) {
  b <- s1[[paste0(cl, "_B")]]; sh <- s1[[paste0(cl, "_S")]]
  data.frame(cell = cl,
             `bag missing` = sum(is.na(b)), `shoe missing` = sum(is.na(sh)),
             `bag M` = round(mean(b, na.rm = TRUE), 3),
             `shoe M` = round(mean(sh, na.rm = TRUE), 3),
             `composite M` = round(mean(s1[[paste0(cl, "_BS")]]), 3),
             check.names = FALSE)
})), row.names = FALSE, caption = "Logo area, square inches.")
```

Bag logos are about three times the size of shoe logos, so substituting the bag value for
the composite inflates a cell. That inflation hits 18 cases in the high-fertility self
cell, 12 in the low-fertility self cell, and none in either other cell. That is the
direction of Hypothesis 1 and of the interaction.

```{r s1-robust}
cc <- complete.cases(s1[, c(outer(cells, c("_B", "_S"), paste0))])
tru <- sapply(cells, function(cl) (s1[[paste0(cl, "_B")]] + s1[[paste0(cl, "_S")]]) / 2)

spec <- function(dat, ses, lab) {
  ix <- (dat[, "High_Self"] - dat[, "High_Other"]) - (dat[, "Low_Self"] - dat[, "Low_Other"])
  h1 <- dat[, "High_Self"] - dat[, "Low_Self"]
  a <- ctest(ix, ses); b <- ctest(h1, ses)
  data.frame(specification = lab, n = a$n,
             `interaction F` = round(a$F, 2), `interaction p` = fmt_p(a$p),
             `H1 F` = round(b$F, 2), `H1 p` = fmt_p(b$p), check.names = FALSE)
}

bs <- as.matrix(s1[, paste0(cells, "_BS")]);  colnames(bs) <- cells
bg <- as.matrix(s1[, paste0(cells, "_B")]);   colnames(bg) <- cells
sh <- as.matrix(s1[, paste0(cells, "_S")]);   colnames(sh) <- cells

kable(rbind(
  spec(bs, s1$SES, "as published (mean of available drawings)"),
  spec(bg, s1$SES, "bags only"),
  spec(sh[complete.cases(sh), ], s1$SES[complete.cases(sh)], "shoes only"),
  spec(tru[cc, ], s1$SES[cc], "complete cases, true 2-item composite")),
  row.names = FALSE, caption = "All covariates correctly centered.")
```

The test of Hypothesis 1 is null in every variant. The interaction survives only in the
bags-only row, where the self cells still differ from the other cells in *n*. In the 51
women who actually produced all eight drawings it is gone (*p* = .22, and *p* = .25
without the covariate). Some of that is lost power from dropping 19 women, but the
published interaction still depends on a composite that is built differently in the two
self cells than in the two other cells.

### Session order is confounded with fertility and fits the interaction at least as well

Order was not counterbalanced, although the paper says in one place that "the order was
counterbalanced" and in another that it "depended on where she was in her menstrual cycle"
at screening, producing 63% high-fertility-first. Order is therefore correlated with
fertility by design, and the two sessions differ in practice/familiarity with the drawing
task. (`Session_Order` is unlabelled in the posted file and takes the values −1 for 44
women and 1 for 26; only 1 = low-fertility-first reproduces the paper's 37%/63% split, so
that is the coding used here.)

```{r s1-order}
with(s1, {
  ix  <- (High_Self_BS - High_Other_BS) - (Low_Self_BS - Low_Other_BS)
  lowfirst <- Session_Order == 1
  tt <- t.test(ix[lowfirst], ix[!lowfirst])

  # recode the same contrast by session number instead of fertility
  self1  <- ifelse(lowfirst, Low_Self_BS,  High_Self_BS)
  self2  <- ifelse(lowfirst, High_Self_BS, Low_Self_BS)
  oth1   <- ifelse(lowfirst, Low_Other_BS, High_Other_BS)
  oth2   <- ifelse(lowfirst, High_Other_BS, Low_Other_BS)
  ix_ses <- (self1 - oth1) - (self2 - oth2)

  kable(data.frame(
    contrast = c("fertility x target, low-fertility-first group (n=26)",
                 "fertility x target, high-fertility-first group (n=44)",
                 "difference between the two groups",
                 "SESSION 1 vs 2 x target (same data, order recoding)",
                 "logos for OTHER women, session 1 vs session 2"),
    M = round(c(mean(ix[lowfirst]), mean(ix[!lowfirst]),
                mean(ix[lowfirst]) - mean(ix[!lowfirst]),
                mean(ix_ses), mean(oth1 - oth2)), 3),
    t = round(c(t.test(ix[lowfirst])$statistic, t.test(ix[!lowfirst])$statistic,
                tt$statistic, t.test(ix_ses)$statistic, t.test(oth1 - oth2)$statistic), 2),
    p = fmt_p(c(t.test(ix[lowfirst])$p.value, t.test(ix[!lowfirst])$p.value,
                tt$p.value, t.test(ix_ses)$p.value, t.test(oth1 - oth2)$p.value))),
    row.names = FALSE)
})
```

The interaction is absent in the 26 women who did the low-fertility session first. The
difference between the two order groups is itself significant (Welch *p* = .017 in the
table; *p* = .036 assuming equal variances). And if you recode the contrast by session
number instead of by fertility, you get an effect at least as strong (*p* = .010) as the
published fertility interaction (*p* = .026). Every woman has exactly one order, so the
design has no way to tell a fertility effect from a session effect. The paper's
Limitations section says the within-subject design "rules out situational differences".
It does not, when the situation (first vs second visit) is confounded with the
predictor.

The last row shows what an order account would look like: women drew larger logos for
other women in their second session than their first, while self-logos did not change. Because 63% of women had their high-fertility session first, a session-2
inflation of other-directed logos would appear in the analysis as "smaller other-logos at
high fertility". This is a between-person comparison, and the two order groups also
differ in where they were in their cycle at screening, so it is an alternative
explanation the design cannot exclude rather than a demonstrated mechanism.

```{r s1-orderfig, fig.cap="Study 1 interaction contrast by session order. Points are participants; diamonds are group means."}
with(s1, {
  ix <- (High_Self_BS - High_Other_BS) - (Low_Self_BS - Low_Other_BS)
  d <- data.frame(ix, order = factor(ifelse(Session_Order == 1, "low fertility first (n=26)",
                                            "high fertility first (n=44)")))
  ggplot(d, aes(order, ix)) +
    geom_hline(yintercept = 0, linetype = 2, colour = "grey50") +
    geom_jitter(width = .12, alpha = .45, size = 1.6) +
    stat_summary(fun = mean, geom = "point", shape = 18, size = 5, colour = "firebrick") +
    labs(x = NULL, y = "(self - other) at high fertility\nminus (self - other) at low fertility")
})
```

### Distributional assumptions

Logo area is bounded at zero, right-skewed, and has SD > M in three of four cells.
The interaction does not survive a change of scale or of test.

```{r s1-dist}
with(s1, {
  ix  <- (High_Self_BS - High_Other_BS) - (Low_Self_BS - Low_Other_BS)
  lix <- (log(High_Self_BS) - log(High_Other_BS)) - (log(Low_Self_BS) - log(Low_Other_BS))
  h1  <- High_Self_BS - Low_Self_BS
  lh1 <- log(High_Self_BS) - log(Low_Self_BS)
  kable(data.frame(
    test = c("interaction, raw t-test", "interaction, Wilcoxon signed rank",
             "interaction, log scale", "H1, raw t-test", "H1, Wilcoxon", "H1, log scale"),
    p = fmt_p(c(t.test(ix)$p.value, wilcox.test(ix)$p.value, t.test(lix)$p.value,
                t.test(h1)$p.value, wilcox.test(h1)$p.value, t.test(lh1)$p.value))),
    row.names = FALSE)
})
```

The skew, and the fact that the paper never says whether the people measuring logos with
a caliper were blind to session, would each deserve a sentence in any published version.

---

## Study 2

### The model reproduces; the conditional effects use the wrong error term

```{r s2-model}
s2$IVc <- s2$IV - mean(s2$IV)
m2 <- lm(DV ~ IVc * BrandType * Setting, data = s2)
kable(round(summary(m2)$coefficients, 3), caption = "PROCESS Model 3, reproduced as OLS.")
```

The three-way interaction reproduces exactly (*b* = 6.53, SE = 2.68, *p* = .015). So do
the paper's conditional effects, including their standard errors (5.47 and 5.87), which
confirms they come from the pooled residual variance of the full model. That pooling
assumes equal error variance across the four cells:

```{r s2-levene}
cell <- interaction(s2$BrandType, s2$Setting)
z <- abs(s2$DV - ave(s2$DV, cell, FUN = median))
lev <- anova(lm(z ~ cell))
cat(sprintf("Brown-Forsythe / Levene: F(%d, %d) = %.2f, p = %.2g\n",
            lev$Df[1], lev$Df[2], lev$`F value`[1], lev$`Pr(>F)`[1]))
kable(data.frame(cell = c("nonluxury / private", "luxury / private",
                          "nonluxury / public",  "luxury / public"),
                 `SD of DV` = round(as.numeric(tapply(s2$DV, cell, sd)), 2),
                 check.names = FALSE), row.names = FALSE)
```

Variances differ by a factor of four (SD 0.90 vs 1.89), and Levene's test rejects
equality at *p* < 10⁻⁵. Pooling borrows precision from the tight nonluxury cells and
lends it to the luxury cells, where the hypotheses live.

```{r s2-slopes}
slope <- function(mod, B, S, V) {
  k <- setNames(rep(0, length(coef(mod))), names(coef(mod)))
  k["IVc"] <- 1; k["IVc:BrandType"] <- B; k["IVc:Setting"] <- S
  k["IVc:BrandType:Setting"] <- B * S
  b <- sum(k * coef(mod)); se <- sqrt(drop(t(k) %*% V %*% k))
  c(b = b, se = se, p = 2 * pt(abs(b / se), mod$df.residual, lower.tail = FALSE))
}
cells2 <- list(c(1, 1), c(1, -1), c(-1, 1), c(-1, -1))
labs2  <- c("luxury / public", "luxury / private", "nonluxury / public", "nonluxury / private")

pooled <- t(sapply(cells2, function(x) slope(m2, x[1], x[2], vcov(m2))))
robust <- t(sapply(cells2, function(x) slope(m2, x[1], x[2], vcovHC(m2, type = "HC3"))))
percell <- t(sapply(cells2, function(x) {
  d <- s2[s2$BrandType == x[1] & s2$Setting == x[2], ]
  cf <- unname(summary(lm(DV ~ IVc, data = d))$coefficients[2, ])
  c(b = cf[1], se = cf[2], p = cf[4])
}))
spear <- sapply(cells2, function(x) {
  d <- s2[s2$BrandType == x[1] & s2$Setting == x[2], ]
  cor.test(d$IV, d$DV, method = "spearman", exact = FALSE)$p.value
})

kable(data.frame(
  `conditional slope` = labs2,
  b = round(pooled[, "b"], 2),
  `p (pooled MSE, as published)` = fmt_p(pooled[, "p"]),
  `p (HC3 robust)` = fmt_p(robust[, "p"]),
  `p (within-cell OLS)` = fmt_p(percell[, "p"]),
  `p (Spearman)` = fmt_p(spear),
  check.names = FALSE), row.names = FALSE)

V3 <- vcovHC(m2, type = "HC3")
ixs <- c("IVc:Setting", "IVc:BrandType:Setting")
kable(data.frame(term = ixs, b = round(coef(m2)[ixs], 2),
                 `p (pooled)` = fmt_p(summary(m2)$coefficients[ixs, 4]),
                 `p (HC3)` = fmt_p(2 * pt(abs(coef(m2)[ixs] / sqrt(diag(V3)[ixs])), m2$df.residual, lower.tail = FALSE)),
                 check.names = FALSE), row.names = FALSE,
      caption = "The interaction terms survive robust standard errors; the simple slopes do not.")
```

The interactions themselves are not in doubt. With HC3 standard errors the three-way
interaction is *p* = .010 and the fertility × setting interaction within luxury brands is
*p* = .005. What changes is the decomposition the paper's conclusions are built on. The
effect that supports Hypothesis 1, fertility raising purchase intention for publicly worn
luxury, is *p* ≈ .08 with robust or within-cell standard errors, and *p* = .19 by rank
correlation. The paper's sentence "Fertility had a positive effect on purchase intention
in a public setting (b = 12.08, SE = 5.47, p = .028)" rests on a standard error that is
too small by about a quarter. The only slope that gets stronger under HC3 correction is
the negative one in the private condition. That contradicts H3 as written, which
predicted no effect in private, not a reversal. And that slope is fragile too: *p* = .07
by within-cell OLS, and gone after removing five influential cases.

```{r s2-influence}
dpriv <- s2[s2$BrandType == 1 & s2$Setting == -1, ]
mp <- lm(DV ~ IV, data = dpriv)
drop5 <- dpriv[-order(cooks.distance(mp), decreasing = TRUE)[1:5], ]
cf <- summary(lm(DV ~ IV, data = drop5))$coefficients[2, ]
cat(sprintf("private luxury, dropping 5 most influential cases: b = %.2f, p = %.3f (n = %d)\n",
            cf[1], cf[4], nrow(drop5)))
```

### The private-luxury dependent variable is mostly floor

```{r s2-floor}
lux <- c("Chanel", "Prada", "Gucci", "LV", "Fendi", "Burberry")
pl <- s2[s2$Setting == -1 & s2$BrandType == 1, paste0("Private_", lux)]
pu <- s2[s2$Setting ==  1 & s2$BrandType == 1, paste0("Public_",  lux)]
alpha <- function(X) { X <- na.omit(X); k <- ncol(X)
  k / (k - 1) * (1 - sum(apply(X, 2, var)) / var(rowSums(X))) }
kable(data.frame(
  block = c("private luxury", "public luxury"),
  `alpha (reported)` = c(".99", ".97"),
  `alpha (computed)` = round(c(alpha(pl), alpha(pu)), 3),
  `% identical across all 6 brands` = round(100 * c(mean(apply(pl, 1, function(r) length(unique(r)) == 1)),
                                                    mean(apply(pu, 1, function(r) length(unique(r)) == 1))), 1),
  `% answering 1 to all 6` = round(100 * c(mean(apply(pl, 1, function(r) all(r == 1))),
                                           mean(apply(pu, 1, function(r) all(r == 1)))), 1),
  check.names = FALSE), row.names = FALSE)
dpriv <- s2[s2$BrandType == 1 & s2$Setting == -1, ]
cf <- summary(glm(I(DV > 1) ~ IV, data = dpriv, family = binomial))$coefficients[2, ]
cat(sprintf("private luxury, logistic regression of any interest (DV > 1) on fertility: b = %.1f, p = %.2f\n", cf[1], cf[4]))
```

Half the private-luxury respondents answered "not at all" to all six brands and 60% gave
six identical answers. The α = .99 the paper reports as evidence of scale quality mostly
measures that people did not differentiate between Chanel and Burberry. Fitting OLS to
that variable and interpreting a slope on it is not sound. An ordinal or censored model is
the minimum, and even then there is very little signal to model. A logistic regression of
"any interest at all" on fertility in this cell gives *p* = .18.

The predictor has the same problem in milder form. Conception probability is bounded at
zero and right-skewed: a third of the sample sits at or below .005 and 20 women sit at
exactly zero, so the "−1 SD" point at which the paper reports its "low fertility" means
is effectively the floor of the scale.

```{r s2-range}
mu <- mean(s2$IV); sdv <- sd(s2$IV)
cat(sprintf("mean = %.4f, SD = %.4f;  -1 SD = %+.4f;  observed min = %.3f;  %% at IV <= .005 = %.1f;  n at 0 = %d\n",
            mu, sdv, mu - sdv, min(s2$IV), 100 * mean(s2$IV <= .005), sum(s2$IV == 0)))
```

### Exclusions

Of 1,486 women recruited, 938 passed the health and contraception screening and 332 were
analysed, a 22% retention rate. The second step alone discards 65% on cycle-date
certainty and cycle length. That is very high, and the paper does not ask why. The
criteria are joined with "and" in Study 2 and with "or" in Study 3, and whether all four
separate certainty ratings had to clear 6 is not stated.

This is not neutral data cleaning. Cycle regularity and self-reported cycle knowledge
depend on hormonal health, age, BMI, stress and conscientiousness. Conditioning the sample
on them, while the exposure of interest is computed from the same self-reports, is a
selection problem that could bias the fertility–outcome association in either direction.
The paper's stated target was *n* = 500. It finished at 332 without comment.

```{r s2-age}
cat(sprintf("age: M = %.2f, SD = %.2f, range %d-%d; n over 40 = %d\n",
            mean(s2$Age), sd(s2$Age), min(s2$Age), max(s2$Age), sum(s2$Age > 40)))
```

Six women over 40 (the oldest is 64) are kept in a conception-probability analysis,
while Study 3 prescreened to 18--40. The two studies disagree with each other about
whether this matters.

---

## Study 3

### The preregistered analysis failed

```{r s3-model}
s3$IVc <- s3$IV - mean(s3$IV)
m3 <- glm(DV ~ IVc * Belief, data = s3, family = binomial)
kable(round(summary(m3)$coefficients, 3))

ss <- function(bel) {
  k <- c(0, 1, 0, bel)
  b <- sum(k * coef(m3)); se <- sqrt(drop(t(k) %*% vcov(m3) %*% k))
  c(b = b, se = se, p = 2 * pnorm(abs(b / se), lower.tail = FALSE))
}
kable(data.frame(
  condition = c("control", "weak belief"),
  b  = round(c(ss(-1)["b"],  ss(1)["b"]), 2),
  SE = round(c(ss(-1)["se"], ss(1)["se"]), 2),
  p  = fmt_p(c(ss(-1)["p"],  ss(1)["p"]))), row.names = FALSE,
  caption = "Simple slopes of fertility on luxury choice.")
```

This reproduces the paper exactly: interaction *b* = −5.83, SE = 3.54, *p* = .099;
control-condition slope *b* = 10.22, SE = 5.56, *p* = .066. The preregistered prediction
was not confirmed.

### The rescue analysis contradicts the preregistration

AsPredicted #153,708 (registered 2023-12-03, three days before data collection began),
section 6, states the analysis exclusions:

> We will only include women with regular ovulation cycles (i.e., between 25 and 36 days,
> inclusive) … Moreover, we will only include women who are relatively certain (i.e., 6 or
> higher on a 9-point scale) about their reported period dates and usual cycle length.

The paper then reports an analysis on *n* = 888 "without excluding women who were unsure
about their menstrual cycle", *p* = .024, and presents it as support. This drops a
preregistered exclusion criterion after the preregistered analysis failed, and the
justification given in the text is the failure itself ("Given the marginally significant
interaction…"). The same paper says that "all analyses, sample size, and exclusion
criteria were pre-registered." Both cannot be true.

Further undisclosed deviations:

| Preregistered | Reported |
|---|---|
| *N* = 1,000 after prescreening | 966 collected |
| Exclusions: cycle length 25–36, certainty ≥ 6 | plus 88 excluded on an unregistered manipulation-check item |
| Regression with fertility × condition | same, but the conclusion rests on the unregistered full sample |

```{r s3-balance}
tb <- table(s3$Belief)
bt <- binom.test(as.integer(tb["1"]), sum(tb), .5)
cat(sprintf("control n = %d, weak-belief n = %d, binomial p = %.4f\n",
            tb["-1"], tb["1"], bt$p.value))
cat(sprintf("manipulation-check variable in posted data: %d unique value(s) -> only passers are posted\n",
            length(unique(s3$mc))))
```

The condition sizes are unbalanced (156 vs 202, *p* = .017). Random assignment should
not produce this; some exclusion step did. One candidate is the unregistered
manipulation-check filter: the control-condition memory question (2 trillion vs 1
trillion galaxies) is harder than the weak-belief one (beauty / smile / walk vs height /
table manners / brand), so it plausibly removed more controls. This cannot be checked,
because neither the 966 nor the 888 cases are posted. But if it is what happened, the
control group is differentially enriched for attentive responders, and attentiveness
plausibly governs both cycle-date certainty (the second exclusion step) and care in
choosing between gift cards. That would break randomisation on exactly the dimension that
matters for the one "significant" simple slope.

The posted file contains only the 358 analysed cases (`mc` is constant), so the
*n* = 888 analysis on which the conclusion depends cannot be checked by anyone.

### What is actually there

```{r s3-tertiles}
s3$tert <- ave(s3$IV, s3$Belief, FUN = function(x) as.integer(cut(x, quantile(x, 0:3/3), include.lowest = TRUE)))
tab <- aggregate(DV ~ Belief + tert, s3, function(x) c(n = length(x), pct = 100 * mean(x)))
tab <- data.frame(condition = ifelse(tab$Belief == -1, "control", "weak belief"),
                  `fertility tertile` = c("low", "mid", "high")[tab$tert],
                  n = tab$DV[, "n"], `% luxury` = round(tab$DV[, "pct"], 1), check.names = FALSE)
kable(tab[order(tab$condition, match(tab$`fertility tertile`, c("low","mid","high"))), ], row.names = FALSE)
```

```{r s3-median}
mcut <- median(s3$IV)
for (b in c(-1, 1)) {
  d <- s3[s3$Belief == b, ]
  ct <- chisq.test(table(d$IV > mcut, d$DV), correct = FALSE)
  cat(sprintf("%-12s low-fert %.1f%% (n=%d) vs high-fert %.1f%% (n=%d), chi2 p = %.3f\n",
              ifelse(b == -1, "control", "weak belief"),
              100 * mean(d$DV[d$IV <= mcut]), sum(d$IV <= mcut),
              100 * mean(d$DV[d$IV >  mcut]), sum(d$IV >  mcut), ct$p.value))
}
```

The control-condition pattern is at least monotone, and the median split is significant.
That is the most credible signal in the whole paper: one marginal moderation on a binary
outcome, which becomes significant only after an unregistered change to the sample, in a
2023–24 MTurk sample prescreened by self-report for recent luxury purchasing.

---

## Fertility measurement

Studies 2 and 3 count backward from the participant's *expected* next period, which is
never checked against an actual onset. This is the weakest of the counting methods. The
"expected" date is itself a forward projection from self-reported cycle length, so it
inherits exactly the error that backward counting is supposed to remove. Gangestad et al.
(2016), the paper's own methodological authority, recommend verifying against the next
onset. Study 1 did that with a post-study phone call, so the authors know the step
exists. Studies 2 and 3 skip it.

Two things follow from this that the paper never deals with.

First, the conception-probability variable is not a hormone. It is computed from days
since reported last onset and reported cycle length, so it correlates with everything
else that tracks cycle day. In a cross-sectional design, "low fertility" pools women who
are menstruating with women who are premenstrual, and both groups have obvious reasons
of their own to be less interested in buying a designer T-shirt to wear in public. A
fertility effect and a "neither bleeding nor premenstrual" effect are never separated.
Modelling cycle day non-linearly, or contrasting matched pre- and post-ovulatory windows,
would have done it.

Second, the estrogen mechanism is asserted throughout and measured nowhere. The
theoretical frame is that "estrogen levels increase and induce a nonconscious mating
goal". Study 1 detects an LH surge, which is a reasonable phase marker but says nothing
about estradiol levels. Studies 2 and 3 measure nothing at all.

Here the methodological literature cuts in a direction the paper does not expect. Arslan
et al. (2023, *Psychoneuroendocrinology* 149: 105994) show that adding salivary estradiol
immunoassays would not have helped. Across more than 1,200 participants and 9,500 time
points, salivary immunoassay estradiol was only weakly predictable from cycle phase and
biased upward relative to serum, and imputing the population-average serum change from a
good phase measure was more valid than assaying the individual. So the right response to
"no hormones were measured" is not that the authors should have collected saliva. It is
that the whole inferential chain rests on the phase measure being good enough to carry
the imputation, and reverse counting from an unverified expected onset is not. The
reference list does contain Arslan et al. (2021) and Schleifenbaum et al. (2021), but
only for substantive claims about desire and self-perceived attractiveness, and none of
the methodological content of those papers shows up.

Both sample-size justifications are for "medium" effects. Large diary and hormone studies
of the last decade put credible ovulatory shifts in the small range for most outcomes,
and well below that for something as downstream as consumer behaviour. When the true
effect is small, powering for *d* = 0.5 means that whatever clears *p* < .05 is inflated
in expectation, which is what the re-analysis shows.

---

## Theory and contribution

The novelty claim is contradicted by the paper's own reference list. "This research
shows a novel factor – high fertility – that has not been previously explored as an
influence on luxury consumption" sits alongside citations to Durante et al. (2011) on
ovulation and product choice, Durante et al. (2014) on money, status and the ovulatory
cycle, Hwang & Zhang (2025) on female conspicuous consumption and mating goals, and Chen
et al. (2023) on luxury as a mate-screening signal.

The stated theoretical contribution is untested. No study manipulates or measures which
phase of the mating process a participant is in, so the distinction between an
attraction phase and a selection phase is a way of reconciling contradictory findings,
not a hypothesis. As used here it absorbs either sign of result. A positive
fertility–luxury association is the attraction phase, a negative one is the selection
phase.

The mediator is never measured. "Being noticed" is the mechanism in every hypothesis and
appears in no dependent variable.

The moderator hypotheses predict null effects. H2, H3 and H4 all predict the *absence* of
an effect, are tested with ordinary significance tests, and are reported as supported
when *p* > .05. There are no equivalence tests and no Bayes factors. Worse, two of them
did not come out null but reversed (the other-women condition in Study 1 and the private
condition in Study 2), and both reversals were reinterpreted after the fact as further
support.

The citation practice is selective in a particular way. Durante, Arsena & Griskevicius
(2014) appears as a methods source with no indication that it is a reply to a critique,
and none of the methodological debate of the past decade appears at all.
Galindo-Caballero et al.'s p-curve is invoked as showing "strong evidential value" for
the ovulatory dress effect, which is a weak defence in a literature where the selection
process that generated the *p*-values is the very thing in dispute.

---

## Transparency of the deposit

The OSF deposit does not support the claims made from it.

- Study 1: `Low_Self_B` is absent from the file, and the shoe-drawing missingness is
  documented neither in the file nor in the paper.
- Study 2: four of the 24 brand-rating items (`Public_Gap`, `Public_OldNavy`,
  `Public_JCPenney`, `Public_HM`) are absent, so the reported public-nonluxury α = .65
  cannot be checked.
- Studies 2 and 3: no cycle dates, no cycle-length or certainty ratings, only the derived
  conception probability. The exclusion pipeline, which is the single largest researcher
  degree of freedom in both studies, cannot be audited.
- Study 3: only the 358 analysed cases, so the *n* = 888 analysis cannot be reproduced.
- No analysis code for any study.

```{r deposit}
kable(data.frame(
  file = c("Study 1", "Study 2", "Study 3"),
  `rows posted` = c(nrow(s1), nrow(s2), nrow(s3)),
  `N in paper` = c("70", "332 (of 1,486 recruited)", "358 (and 888 for the key analysis)"),
  `columns as posted` = posted_ncol, check.names = FALSE), row.names = FALSE)
```

Study 1's data were collected between September 2013 and July 2014 and published in
2026. The paper does not say what else was measured in those two lab sessions.

---

## Smaller errors

These matter mainly as an indication of how carefully the paper was written, and of what
peer review caught.

- Study 1 pretest: *M* = 7.82, *SD* = 1.16, tested against the scale midpoint of 5,
  reported as Cohen's *d* = 1.16. The correct value is (7.82 − 5) / 1.16 = 2.43. The SD
  seems to have been copied into the *d* slot.
- Pretest degrees of freedom are internally inconsistent: *F*(1, 75) and *F*(1, 76) from
  the same *n* = 78 pretest; *F*(1, 66) and *F*(1, 65) from the same *n* = 67 pretest.
- "(Otterbring et al., 2020)" appears in author–date format in the Future Research
  section and is absent from the numbered reference list (ref. 48 is Otterbring et al.
  2018).
- Figure 3's y-axis reads "% Chocie for Luxury Brand".
- "yatch", "athough", "pratices", "subects", "parter", "η² = 01", "*p* >= .421".
- Study 3 reports the same IRB number as Study 2 (Pro2020000942, approved 2020) for data
  collected in 2023–24. This may be a continuing protocol, but it is not explained.

---


## Citations that do not support the claims attached to them

All 60 references exist. Checked against Crossref, the author lists, journals, volumes and
page ranges are essentially all correct, and the DOIs and PMIDs resolve to the right
papers. The problem is one level up. Several claims are attributed to sources that do not
contain them, and the two citations that carry the paper's methodological defence both say
close to the opposite of what they are cited for.

### The fertility-measurement warrant is reversed

Reference 36, Gangestad, Haselton, Welling, Gildersleeve, Pillsworth, Burriss, Larson &
Puts (2016), is cited ten times and underwrites almost every methodological decision in the
paper. Its finding, from simulating over 58,000 cycles against published follicular and
luteal phase-length distributions, is that aside from methods counting backward from a
*confirmed* onset of next menses, the validity of counting methods is modest, generally in
the region of .40 to .55.

Studies 2 and 3 say they use "reverse cycle counting", and the only next-onset date they
collect is the one the participant *anticipates*, so the count must run from that. That is
one of the methods the cited paper evaluates and finds wanting; only the confirmed-onset
version survives. Yet the text reads:

> Previous research [36] supports this as the most reliable method to estimate fertility
> based on cross-sectional data points.

This reverses the source. Study 1 verified the next onset with a post-study phone call, so
the authors knew the distinction. It just was not applied in the two studies that lean on
[36] to justify skipping it.

Reference 36 is also cited for two sample-size prescriptions, *n* = 50 for within-subject
LH-confirmed designs and *n* = 500 for cross-sectional designs, both for medium effects.
The paper does report power estimates, so these figures may be in it. But any such
estimate is conditional on the validity of the measurement method, and the validity
reported for anticipated-onset counting is .40–.55. Citing a power target derived for one
method to justify a sample size while using a weaker method is at best incomplete.

### The evidential-value warrant is reversed

Reference 20, Galindo-Caballero, Alzate-Pamplona, Gangestad & Cruz (2023), appears in the
text as "a recent meta-analysis" showing that the ovulatory dress effect "has strong
evidential value". It is a narrative review with a p-curve analysis, not a meta-analysis,
and its own conclusions are:

- evidence exists for peri-ovulatory associations with specific consumer preferences,
  most strongly for appearance-enhancing products;
- there is no compelling evidence for associations with financial decisions and
  risk-taking;
- the authors close by calling for studies "that possess sufficient statistical power to
  detect true meaningful effects".

Luxury purchase intentions and a choice between two $100 gift cards are not
appearance-enhancing products. They sit nearer the financial-decision category the review
says lacks support. The call for better-powered studies is dropped entirely by a paper
that presents final samples of 332 and 358 as adequate.

Reference 20 is cited a second time to support "Near ovulation, women may behave in different
ways with the goal of being noticed [20]". A p-curve over published *p*-values cannot speak
to what goal produced a behaviour.

Gangestad is an author on both [20] and [36]. The paper cites him twice as its
methodological authority and contradicts him both times.

### Other citation–claim mismatches

| Claim in the text | Cited as | Problem |
|---|---|---|
| "Consumers spend billions of dollars each year on publicly visible high-end goods such as designer clothes and accessories, jewelry, and luxury cars" | [1] Angioni & Donini 2017 | A seven-page qualitative comparative case study of the Italian Tod's Group in a university economics annals. Contains no aggregate spending figures. |
| "women account for nearly half of the spending on luxury goods and services" | [5] Statista | The underlying dataset (Start.io, an ad-tech firm) is the distribution of US *mobile-device* luxury shoppers by age and gender. Shoppers are not spending, mobile-app users are not the luxury market, and the same source's UK and Germany figures show men far outnumbering women, so it may not support "nearly half" in any reading. |
| ovulation may lead women to engage in behaviours such as "telling stories and jokes that would receive attention [22–24]" | [24] Greengross & Miller 2011 | A study of humour ability as a fitness indicator, whose finding is that humour ability is higher in males. Nothing about ovulation or about women's attention-seeking. |
| "women near ovulation feel more attractive and socialize more [14–16]" | [16] Ulu, Durante, Berger & Kim 2018 | A two-page *Advances in Consumer Research* conference abstract on word-of-mouth behaviour, not a peer-reviewed report, co-authored by this paper's first and second authors. |
| "motivations other than mating may shift women's luxury consumption [60]" | [60] Young-Eisendrath 2023 | A 1999 Jungian trade book reissued by a small analytical-psychology press in 2023, cited as though it were an empirical source; the reference gives no indication of what it is. |
| "we used a paradigm adopted in research showing that consumers' desire for conspicuous products is associated with a preference for larger luxury brand logos [30,40,41]" | [30] Lee & Shrum 2012; [40] Nunes et al. 2011; [41] Wang & Griskevicius 2014 | All three measured *choice among or ratings of* products with differently prominent logos. None had participants draw logos freehand and measured the drawn area with a caliper. The Study 1 measure is new and is presented as established; no reliability or validity evidence is offered for it. |
| "ovulation may lead women to engage in exaggerated behaviors that could draw attention to them, such as wearing brighter colors" | [22] Beall & Tracy 2013 | Presented as settled. The red/pink-at-peak-fertility finding has a contested replication record, and nothing in the text signals this. |
| screening procedures follow "the screening materials presented in previous research [39]" | [39] Durante, Arsena & Griskevicius 2014 | Cited purely as a source of screening materials. It is a published *reply to a critique* of the authors' own earlier ovulatory-cycle work; readers are given no indication that the cited exchange exists. |
| "they experience increased sexual desire [13]"; "women feel more attractive before ovulation [15]" | [13] Arslan et al. 2021; [15] Schleifenbaum et al. 2021 | Used only for substantive conclusions. These are the large, well-measured studies whose methodological apparatus bears directly on Studies 2 and 3, and none of that content appears anywhere in the paper. |


## Conclusion

An accurate report of these data would read roughly like this. In a within-subject,
LH-confirmed design, fertility had no effect on self-directed conspicuous luxury
(*p* = .18). The one interaction present is explained at least as well by session order
as by cycle phase, and it vanishes in complete cases. In a cross-sectional design, the
interaction held up, but the slope supporting the hypothesis was *p* ≈ .08 with correct
standard errors, while the only slope that stayed significant ran the wrong way, on a
dependent variable that was half floor. In a preregistered study, the preregistered test
failed at *p* = .099 and was replaced by an analysis that dropped a preregistered
exclusion and cannot be reproduced from the posted data.

That is a null-to-ambiguous set of results, which is what large, well-measured cycle
studies have been finding. The published version differs from it by an uncentered
covariate, a pooled error term, an unregistered sample expansion, and a post-hoc
two-phase theory flexible enough to absorb the reversals.

The citations point the same way. The two references that carry the paper's
methodological defence say close to the opposite of what they are cited for. Gangestad et
al. (2016) find that counting methods other than backward counting from a *confirmed*
next onset have validities of only .40–.55, and are cited as endorsing counting from an
anticipated onset as "the most reliable method". Galindo-Caballero et al. (2023) find
evidence mainly for appearance-enhancing products, none for financial decisions, and call
for better-powered studies, and are cited as a "meta-analysis" showing "strong
evidential value". Several further claims are attached to sources that do not contain
them. A reader who checked the citations would have been warned off the design. A reader
who trusted them was misled about the state of the literature.

None of this individually implies bad faith. The SPSS covariate behaviour in particular
is a real and widespread trap, and not something a reviewer would catch from a results
paragraph. But the errors all run one way, and every one of them could have been caught
from the deposited data before publication. The Study 1 issue alone removes the paper's
support for its title claim in the one hormonally confirmed study, and anyone can verify
it in a few lines of code. Together with the Study 2 error term and the Study 3 deviation
from the preregistration, none of the three studies supports the title as published. At
minimum this calls for a correction that revises the conclusions. Whether the errors,
combined with design problems that no reanalysis can fix, cross into retraction is a
judgment call for the journal.

```{r session}
sessionInfo()
```